---
title: "Checking data against expectations in a data preparation pipeline with expectdata"
author: "Dan Garmat"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Checking data against expectations in a data preparation pipeline with expectdata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Expectdata is an R package that makes it easy to check assumptions about a data frame before conducting analyses. Below is a concise tour of some of the things expectdata can do for you.

## Check for unexpected duplication

```{r, error = TRUE, purl = FALSE}
library(expectdata)
expect_no_duplicates(mtcars, "cyl")
```

The default `return_df == TRUE` option allows for using these function as part of a dplyr piped expression that is stopped when data assumptions are not kept.

```{r}
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
mtcars %>% 
  filter(cyl == 4) %>% 
  expect_no_duplicates("wt", return_df = TRUE) %>% 
  ggplot(aes(x = wt, y = hp, color = mpg, size = mpg)) +
  geom_point()
```

If there are no expectations violated, an "OK" message is printed.

After joining two data sets you may want to verify that no unintended duplication occurred. Expectdata allows comparing pre- and post- processing to ensure they have the same number of rows before continuing.

```{r}
expect_same_number_of_rows(mtcars, mtcars, return_df = FALSE)
expect_same_number_of_rows(mtcars, iris, show_fails = FALSE, stop_if_fail = FALSE, return_df = FALSE)

# can also compare to no df2 to check is zero rows
expect_same_number_of_rows(mtcars, show_fails = FALSE, stop_if_fail = FALSE, return_df = FALSE) 
```

Can see how the `stop_if_fail = FALSE` option will turn failed expectations into warnings instead of errors. 

## Check for existance of problematic rows

Comparing a data frame to an empty, zero-length data frame can also be done more explicitly. If the expectations fail, cases can be shown to begin the next step of exploring why these showed up.

```{r, error = TRUE, purl = FALSE}
expect_zero_rows(mtcars[mtcars$cyl == 0, ], return_df = TRUE)
expect_zero_rows(mtcars$cyl[mtcars$cyl == 0])
expect_zero_rows(mtcars, show_fails = TRUE)
```

This works well at the end of a pipeline that starts with a data frame, runs some logic to filter to cases that should not exist, then runs `expect_zero_rows()` to check no cases exist.

```{r, error = TRUE, purl = FALSE}
# verify no cars have zero cylindars
mtcars %>% 
  filter(cyl == 0) %>% 
  expect_zero_rows(return_df = FALSE)
```

Can also check for NAs in a vector, specific columns of a data frame, or a whole data frame.

```{r, error = TRUE, purl = FALSE}
expect_no_nas(mtcars, "cyl", return_df = FALSE)
expect_no_nas(mtcars, return_df = FALSE)
expect_no_nas(c(0, 3, 4, 5))
expect_no_nas(c(0, 3, NA, 5))
```
